\dnest~Manual

### 2016-01-14 doc/manual.tex
Fixed some problems

The plots are useful for diagnosing the progress of your run. For example, you
want to ensure that the Metropolis acc
eptance fraction doesn't drop to near
zero. If it does, the proposal distributions might be too aggressive, or maybe
you just created too many levels (and the high level distributions are just
extremely narrow). The other key plot for diagnostics is the posterior
weights. On most problems, these will peak at some point, and begin to
come back down towards zero (See Figure~\ref{fig3}).
Once this occurs, you probably have enough levels.
It is wasteful to create large amounts of levels extending well beyond the peak.
However, on problems with phase transitions more than one peak may occur.
An example of not having enough levels is given in
Figure~\ref{fig:not_enough_levels}. To change the number of levels, you
can use the OPTIONS file, described in Section~\ref{sec:options}.\\

@@

\section{Implementing models}
\dnest~currently includes three example models: SpikeSlab, StraightLine, and
RJObject_SineWaves. It also contains an empty Template example which you can use
to implement your own models.\\

To use \dnest~on a particular problem, you need to implement a C++ class
which describes the features of the model (e.g. what the parameters are,
what the prior is, and what the likelihood is). Specifically, an object
(instance) of your class represents a point in the parameter space of your
problem. The class must define several member functions which can be
applied to any object. The first is {\tt from_prior(DNest4::RNG&)},
which when called,
initialises the object with parameter values from the prior distribution
(the argument is a random number generator object, passed by reference).
The second important member function is {\tt perturb(DNest4::RNG&)}, which is used to propose
a change to the parameters, as in the basic Metropolis algorithm. The third
is {\tt log_likelihood()}, which returns (as you might expect)
the log of the likelihood function evaluated at the object's current position. Next,
we have the {\tt print(std::ostream\&)} member function which defines how the model
parameters are to be written to the file (i.e. which parameters are to be
printed, and in what order). Finally, there is {\tt description()},
which allows you to put a header at the top of the output files with information
about what is in each column. To demonstrate how these member functions work, we'll go
through the code for {\tt SpikeSlab}.\\

@@

This constructor does nothing except specify the size of the {\tt params}
valarray, which in this example is 20. Next we have {\tt from_prior(DNest4::RNG&)}, the
function which generates the 20 parameters from the prior. In the SpikeSlab
example, the prior is Uniform$(-0.5, 0.5)$, so this function just has a loop
which generates each value from this distribution. Here I'm making use of
a function {\tt rng.rand()} which generates values from $U(0, 1)$ using the
random number generator called rng (which is passed in by \dnest).
The RNG class is defined in {\tt RNG.h} within \dnest, and is really just a
wrapper of C++11's random number generator, with more convenient function names.

@@

The next important member function is {\tt perturb(DNest4::RNG&}, which defines the Metropolis
proposals that are used to explore. Since \dnest~spends some time sampling
the prior, and some time sampling narrow distributions like the posterior,
it's a good idea to use very heavy tailed proposal distributions. In the
SpikeSlab example, the prior width is 1, so you shouldn't make moves much
bigger than that, but you don't have any real idea how narrow the posterior
distribution is until you've solved the problem. In the SpikeSlab example,
the proposal is done by choosing one of the 20 parameters to move, and making
a move according to a scale mixture of gaussians (equivalent to ``a gaussian
proposal with a random step size''). Finally, to prevent any proposals outside
the prior range, I use modulo to ensure the proposed value is between -0.5
and 0.5. The resulting proposal is:\\

@@

The {\tt log_likelihood} member does what you'd expect: evaluates and returns
the log likelihood based on the current setting of the parameters. In the
SpikeSlab example, the likelihood is an invented mixture of two gaussians.
However, in general, your likelihood will depend on a data set. When I'm using
real data, I usually write a {\tt Data} class, an object of which represents
a data set. Since you need to access the dataset from within another class,
it's easiest to make the C++ equivalent of a ``global variable'': a single
static instance of the dataset which is accessible from everywhere. See
the StraightLine example if you want to see a Data class.\\

@@

The Makefile in the Template directory should allow you to compile your model
by issuing the command {\tt make} in a terminal. The only preparation you need
to do is add the path to your \dnest~directory to an environment variable called
DNEST4\_PATH so that the compiler knows where \dnest~is on your system.
For example, on my laptop, \dnest~is located in /home/brewer/Projects/DNest4.
So I set DNEST4_PATH to /home/brewer/Projects.

>>>>

The plots are useful for diagnosing the progress of your run. For example, you
want to ensure that the Metropolis acc
eptance fraction doesn't drop to near
zero. If it does, the proposal distributions might be too aggressive, or maybe
you just created too many levels (and the high level distributions are just
extremely narrow). The other key plot for diagnostics is the posterior
weights. On most problems, these will peak at some point, and begin to
come back down towards zero (See Figure~\ref{fig:fig3}).
Once this occurs, you probably have enough levels.
It is wasteful to create large amounts of levels extending well beyond the peak.
However, on problems with phase transitions more than one peak may occur.
An example of not having enough levels is given in
Figure~\ref{fig:not_enough_levels}. To change the number of levels, you
can use the OPTIONS file, described in Section~\ref{sec:options}.\\

@@

\section{Implementing models}
\dnest~currently includes three example models: SpikeSlab, StraightLine, and
RJObject\_SineWaves. It also contains an empty Template example which you can use
to implement your own models.\\

To use \dnest~on a particular problem, you need to implement a C++ class
which describes the features of the model (e.g. what the parameters are,
what the prior is, and what the likelihood is). Specifically, an object
(instance) of your class represents a point in the parameter space of your
problem. The class must define several member functions which can be
applied to any object. The first is {\tt from\_prior(DNest4::RNG\&)},
which when called,
initialises the object with parameter values from the prior distribution
(the argument is a random number generator object, passed by reference).
The second important member function is {\tt perturb(DNest4::RNG\&)}, which is used to propose
a change to the parameters, as in the basic Metropolis algorithm. The third
is {\tt log\_likelihood()}, which returns (as you might expect)
the log of the likelihood function evaluated at the object's current position. Next,
we have the {\tt print(std::ostream\&)} member function which defines how the model
parameters are to be written to the file (i.e. which parameters are to be
printed, and in what order). Finally, there is {\tt description()},
which allows you to put a header at the top of the output files with information
about what is in each column. To demonstrate how these member functions work, we'll go
through the code for {\tt SpikeSlab}.\\

@@

This constructor does nothing except specify the size of the {\tt params}
valarray, which in this example is 20. Next we have {\tt from\_prior(DNest4::RNG\&)}, the
function which generates the 20 parameters from the prior. In the SpikeSlab
example, the prior is Uniform$(-0.5, 0.5)$, so this function just has a loop
which generates each value from this distribution. Here I'm making use of
a function {\tt rng.rand()} which generates values from $U(0, 1)$ using the
random number generator called rng (which is passed in by \dnest).
The RNG class is defined in {\tt RNG.h} within \dnest, and is really just a
wrapper of C++11's random number generator, with more convenient function names.

@@

The next important member function is {\tt perturb(DNest4::RNG\&}, which defines the Metropolis
proposals that are used to explore. Since \dnest~spends some time sampling
the prior, and some time sampling narrow distributions like the posterior,
it's a good idea to use very heavy tailed proposal distributions. In the
SpikeSlab example, the prior width is 1, so you shouldn't make moves much
bigger than that, but you don't have any real idea how narrow the posterior
distribution is until you've solved the problem. In the SpikeSlab example,
the proposal is done by choosing one of the 20 parameters to move, and making
a move according to a scale mixture of gaussians (equivalent to ``a gaussian
proposal with a random step size''). Finally, to prevent any proposals outside
the prior range, I use modulo to ensure the proposed value is between -0.5
and 0.5. The resulting proposal is:\\

@@

The {\tt log\_likelihood} member does what you'd expect: evaluates and returns
the log likelihood based on the current setting of the parameters. In the
SpikeSlab example, the likelihood is an invented mixture of two gaussians.
However, in general, your likelihood will depend on a data set. When I'm using
real data, I usually write a {\tt Data} class, an object of which represents
a data set. Since you need to access the dataset from within another class,
it's easiest to make the C++ equivalent of a ``global variable'': a single
static instance of the dataset which is accessible from everywhere. See
the StraightLine example if you want to see a Data class.\\

@@

The Makefile in the Template directory should allow you to compile your model
by issuing the command {\tt make} in a terminal. The only preparation you need
to do is add the path to your \dnest~directory to an environment variable called
DNEST4\_PATH so that the compiler knows where \dnest~is on your system.
For example, on my laptop, \dnest~is located in /home/brewer/Projects/DNest4.
So I set DNEST4\_PATH to /home/brewer/Projects.
